Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-16 tex/classical.tex
quantum.tex reformatted: 80chars or less

The $H$-theorem fixes the time evolution direction of systems in nature, such
as the second law of thermodynamics does. {\color{blue}footnote: }The molecular hypothesis is %
fundamental to determine the physical meaning of the $H$ theorem. For more %
information, see \cite{bib:huang}.

>>>>

The $H$-theorem fixes the time evolution direction of systems in nature, such
as the second law of thermodynamics does.\footnote{The molecular hypothesis is %
fundamental to determine the physical meaning of the $H$ theorem. For more %
information, see \cite{bib:huang}.}

### 2020-12-16 tex/quantum.tex
quantum.tex reformatted: 80chars or less

\section{Quantum scheme}
\subsection{H theorem and the Fermi-Dirac and Bose-Einstein distribution}
The $H$ functional in quantum mechanics presented by Tolman \cite{tolman} is defined as 
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an equation for the rate of change in the number of particles in group $\kappa$ is
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where $V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions are obtained by the perturbation theory.\\
\textcolor{red}{Tolman's proposal for the quantum $H$-theorem establishes that if we have a quantum gas that satisfies the rate of change equation (\ref{changen})\footnote{It is important to remark that the equal a priori probabilities and random a priori phases hypothesis are implicit in the rate of change equation (\ref{changen}) and the random a priori phases hypothesis is considered as the quantum analogous to the molecular chaos hypothesis \cite{binary}}, then the $H$ functional (\ref{quantumh}) satisfies that
\begin{equation}
    \frac{dH}{dt}\leq 0,
\end{equation}
and as a consequence of this theorem, when the system is in equilibrium, $\frac{dH}{dt}=0$.\\
The proof of this theorem is in the same way as in the classical version. Also, when the system is in equilibrium, the distribution functions that correspond to the equilibrium state are the Fermi-Dirac and the Bose-Einstein distribution. Those distributions can be obtained from a sufficient condition for the equilibrium state
\begin{equation}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}},
\end{equation}
where the upper sign refers to the Bose-Einstein distribution and the lower sign refers to the Fermi-Dirac distribution.}\\
In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the limit of high-energy levels to
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
This expression also can be obtained also from the Boltzmann $H$ functional through defining 
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
%--------------------------
\subsection{Out of equilibrium, non-homogeneous distributions for degenerated systems}

We begin by defining the form of the functional 
\begin{eqnarray}
    \Ss (t)&=&\sum_{M=1}^{K} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]   \delta V_M \delta \epsilon_n \label{entropy}.
\end{eqnarray}
This functional will be named \textit{variational entropy}.
This functional describes, as in the classical case, a system with a global energy $E$, a global number of free
quantum particles (which may be fermions for upper sing or bosons for lower sing) $N$, global volume $V$, and a temperature $T$. 
Besides, the variational entropy should correspond to a density of entropy when the system arrives at the equilibrium state. The
global system, in turn, consists of a set of $K$ small cells with a constant volume $\delta V_M$, whose local homogeneous distribution functions is $f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position of the cells, the energy per cell $\epsilon_{n}$, and time $t$. Also, the energy spectra will be divided into $\delta \epsilon_n$ volumes. We will use the uppercase Latin index to denote cell numbers, and unless stated otherwise, $M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each cell has its local chemical potential, $\mu_M$, local
variational functional, $\Sm$, local number of particles, $\mathcal{N}_M$, and local energy, $\mathcal{E}_M$. In terms of the local distribution 
function and local energy spectra, $\{\epsilon_{n}\}$, the last three properties are given by:
\begin{eqnarray}
    \Sm &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.
\end{eqnarray}
When the system is in equilibrium, the local number of particles and the local energy don't depend on the cell number, this is
\begin{equation}
   {\mathcal{N}}_M=\mathcal{N}\equiv \bar{\mathcal{N}}; \ \ \ \  {\mathcal{E}}_M=\mathcal{E}\equiv \bar{\mathcal{E}}.
\end{equation}
The particles in the system are considered to be free, \textit{i.e.} the particles do not interact with each other.
Consequently, we can assume that the available quantum states do not depend on the cell features,
therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the system is subjected to the microcanonical restrictions
\begin{eqnarray}
    &&\sum_{M=1}^{K} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M=1}^{K}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}
\end{eqnarray}
where $N$ and $E$ are the total particle number and the total energy of the system respectively.\\ 
Maximizing the variational entropy $\Ss$ using the Lagrange multipliers method, we obtain
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain from the variational procedure that 
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
where we used that the $H$ functional in equilibrium is proportional to the entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the variation of the logarithm with the number of particles by cells is indifferent from a particular cell. With this, we conclude  
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
In equilibrium, the energy and the Lagrange multiplier do not depend on the cell number and time, therefore
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}{}
If one compares those multipliers with the mean energy and the mean particle number from the statistical mechanic's results, we obtain
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}{}
and finally
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}{}
We proved, when the system is in equilibrium, the Lagrange multipliers don't depend on the cell number. In other words, Lagrange multipliers are homogeneous in the system.\\
We can calculate the entropy from an ideal system, in other words, if one substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain
\begin{eqnarray}
      \Ss(t)&=&\sum_{M=1}^{K} \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.
  \end{eqnarray}
  This result allows us to calculate the entropy of a quantum ideal gases. This entropy could be used as a thermodynamic variable to describe those systems.

%\subsection{The quantum analogous of the Boltzmann's $H$ theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
\begin{equation}
    \Ss(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

This distribution function must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells), and the following restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. The previous conditions and restrictions are the quantum analogous case of the classical case. Also, this system will suffer, as in the classical case, a \textit{relaxation process}.\\
  The quantities $\Delta_M,\delta_M$ are sufficiently big to be different from fluctuations in the system but sufficiently small to consider the system not far from the equilibrium state, as in the classical case.  \\
If we perform the derivative of the variational entropy with respect to time, we obtain
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation (\ref{firstorder}), we get
\begin{eqnarray}
    \frac{d\Ss (t)}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\
    \label{cambioH1}
\end{eqnarray}{}
It's necessary to remark that, in the case of fermions
\begin{eqnarray}
   1-\bar f_n -\bar f_n g_{nM}>0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}>1+g_{nM}. \label{fermionrestriction}
\end{eqnarray}{}
The previous expression establishes that the value of $g_{nM}$ is determined by $\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy levels are occupied, then the system will not have inhomogeneities because of the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result corresponds to the system in the zero temperature condition. We can observe that in this condition, (\ref{fermionrestriction}) is violated. Due to this, the zero temperature condition is excluded.\\
\\
We can observe in the case of bosons, if $\bar{f}_n \gg \bar{f}_n |g_{nM}|$ then $1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$. In fermions, in the case of a non-extremely degenerated condition, it is true that $1-\bar{f}_n \gg \bar{f}_n |g_{nM}|$. With those relations, we can approximate the logarithm functions in (\ref{cambioH1}) to their first-order Taylor series around $\bar f_n g_{nM}=0$ 
\begin{equation}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM}; \ \ \ \ \ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}] \approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}. \label{lnapproximation}
\end{equation}{}
With the previous approximation, (\ref{cambioH1}) will be
\begin{eqnarray}
    \frac{d\Ss}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M=1}^{K}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M \delta \epsilon_n.\label{cambioH2}
\end{eqnarray}{}
Making use of the expression (\ref{relation}),
(\ref{cambioH2}) casts into
\begin{eqnarray}
    \frac{d\Ss}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \} \de V_M \delta \epsilon_n. \nonumber \\
    \label{cambioH3}
\end{eqnarray}{}
On the other hand, we obtain from the restrictions the following expressions
\begin{eqnarray}
    &&\sum_n \bar{f}_n g_{nM} \delta \epsilon_n=\Delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM} \delta \epsilon_n=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n  \bar{f}_n g_{nM}\epsilon_n \delta \epsilon_n=\delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n \delta \epsilon_n=\dot{\delta}_M(t)
\end{eqnarray}{}
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) \de V_M  =\sum_{M=1}^{K} \delta_M(t) \de V_M =0$, we find
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t) \de V_M =\sum_{M=1}^{K} \dot{\delta}_{M}(t) \de V_M=0.
\end{equation}{}

>>>>

\section{Quantum scheme}

\subsection{H theorem and the Fermi-Dirac and Bose-Einstein distribution}

The $H$ functional in quantum mechanics presented by Tolman \cite{bib:tolman}
is defined as
%
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
%
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an
equation for the rate of change in the number of particles in group $\kappa$ is
%
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
%
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
%
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
%
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions
are obtained by the perturbation theory.

\textcolor{red}{Tolman's proposal for the quantum $H$-theorem establishes that
if we have a quantum gas that satisfies the rate of change equation
(\ref{changen})\footnote{It is important to remark that the equal a priori
probabilities and random a priori phases hypothesis are implicit in the rate of
change equation (\ref{changen}) and the random a priori phases hypothesis is
considered as the quantum analogous to the molecular chaos hypothesis
\cite{bib:das2018}}, then the $H$ functional (\ref{quantumh}) satisfies that
%
\begin{equation}
    \frac{dH}{dt}\leq 0,
\end{equation}
%
and as a consequence of this theorem, when the system is in equilibrium,
$\frac{dH}{dt}=0$.\\
The proof of this theorem is in the same way as in the classical version. Also,
when the system is in equilibrium, the distribution functions that correspond
to the equilibrium state are the Fermi-Dirac and the Bose-Einstein
distribution. Those distributions can be obtained from a sufficient condition
for the equilibrium state
%
\begin{equation}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}},
\end{equation}
%
where the upper sign refers to the Bose-Einstein distribution and the lower
sign refers to the Fermi-Dirac distribution.}

In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the
limit of high-energy levels to
%
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
%
This expression also can be obtained also from the Boltzmann $H$ functional through defining
%
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
%
%--------------------------

\subsection{Out of equilibrium, non-homogeneous distributions for degenerated systems}

We begin by defining the form of the functional
%
\begin{eqnarray}
    \Ss (t)&=&\sum_{M=1}^{K} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]   \delta V_M \delta \epsilon_n \label{entropy}.
\end{eqnarray}
%
This functional will be named \textit{variational entropy}.
This functional describes, as in the classical case, a system with a global
energy $E$, a global number of free
quantum particles (which may be fermions for upper sing or bosons for lower
sing) $N$, global volume $V$, and a temperature $T$.
Besides, the variational entropy should correspond to a density of entropy when
the system arrives at the equilibrium state. The
global system, in turn, consists of a set of $K$ small cells with a constant
volume $\delta V_M$, whose local homogeneous distribution functions is
$f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position
of the cells, the energy per cell $\epsilon_{n}$, and time $t$. Also, the
energy spectra will be divided into $\delta \epsilon_n$ volumes. We will use
the uppercase Latin index to denote cell numbers, and unless stated otherwise,
$M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each
cell has its local chemical potential, $\mu_M$, local
variational functional, $\Sm$, local number of particles, $\mathcal{N}_M$, and
local energy, $\mathcal{E}_M$. In terms of the local distribution
function and local energy spectra, $\{\epsilon_{n}\}$, the last three
properties are given by:
%
\begin{eqnarray}
    \Sm &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.
\end{eqnarray}
%
When the system is in equilibrium, the local number of particles and the local
energy don't depend on the cell number, this is
%
\begin{equation}
   {\mathcal{N}}_M=\mathcal{N}\equiv \bar{\mathcal{N}}; \ \ \ \  {\mathcal{E}}_M=\mathcal{E}\equiv \bar{\mathcal{E}}.
\end{equation}
%
The particles in the system are considered to be free, \textit{i.e.} the
particles do not interact with each other.
Consequently, we can assume that the available quantum states do not depend on
the cell features,
therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all
cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the
system is subjected to the microcanonical restrictions
%
\begin{eqnarray}
    &&\sum_{M=1}^{K} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M=1}^{K}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}
\end{eqnarray}
%
where $N$ and $E$ are the total particle number and the total energy of the
system respectively.

Maximizing the variational entropy $\Ss$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$ functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

We can calculate the entropy from an ideal system, in other words, if one
substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain
%
\begin{eqnarray}
      \Ss(t)&=&\sum_{M=1}^{K} \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.
  \end{eqnarray}
%
This result allows us to calculate the entropy of a quantum ideal gases. This
entropy could be used as a thermodynamic variable to describe those systems.

%\subsection{The quantum analogous of the Boltzmann's $H$ theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \Ss(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

This distribution function must satisfy the local equilibrium hypothesis in
each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the
non-homogeneous distribution function assumption in the total volume ($f_{nM}$
must be different among all cells), and the following restrictions
%
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
%
where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle
number and the local energy in equilibrium
%
\begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
%
and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$
and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$
and $\delta_M(t) \ll \bar{\mathcal{E}}$. The previous conditions and
restrictions are the quantum analogous case of the classical case. Also, this
system will suffer, as in the classical case, a \textit{relaxation process}.

  The quantities $\Delta_M,\delta_M$ are sufficiently big to be different from
fluctuations in the system but sufficiently small to consider the system not
far from the equilibrium state, as in the classical case.

If we perform the derivative of the variational entropy with respect to time,
we obtain
%
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}
%
Using the first-order approximation (\ref{firstorder}), we get
%
\begin{eqnarray}
    \frac{d\Ss (t)}{dt}&=&\sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\
    \label{cambioH1}
\end{eqnarray}
%
It's necessary to remark that, in the case of fermions
%
\begin{eqnarray}
   1-\bar f_n -\bar f_n g_{nM}>0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}>1+g_{nM}. \label{fermionrestriction}
\end{eqnarray}
%
The previous expression establishes that the value of $g_{nM}$ is determined by
$\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy
levels are occupied, then the system will not have inhomogeneities because of
the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result
corresponds to the system in the zero temperature condition. We can observe
that in this condition, (\ref{fermionrestriction}) is violated. Due to this,
the zero temperature condition is excluded.
